---
title: "data"
author: "Sahir"
date: "January 22, 2017"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(eclust)
```

# Overview of Functions

This package has three sets of functions starting with either `r_`, `s_` or `u_` 

1. `r_` (**real data functions**): related to analysis of real data. Most users will apply this set of functions to their data. 

```{r, echo=FALSE}
pander::pander(data.frame(`function name` = grep("^r_",pacman::p_functions("eclust"), value = T)))
```


2. `s_` (**simulation functions**): related to the simulations conducted in the [paper](http://sahirbhatnagar.com/slides/manuscript1_SB_v4.pdf). There are functions to simulate data, run the analyses on these data, and output performance metrics.

```{r, echo=FALSE}
pander::pander(data.frame(`function name` = grep("^s_",pacman::p_functions("eclust"), value = T)))
```


3. `u_` (**utility functions**): functions that are used by both simulation and real data analysis functions. Not really meant to be called by the user.  

```{r, echo=FALSE}
pander::pander(data.frame(`function name` = grep("^u_",pacman::p_functions("eclust"), value = T)))
```
