---
title: "data"
author: "Sahir"
date: "January 22, 2017"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(eclust)
```

# Real data analysis functions (`r_`)

We will use the `data(tcgaov)` dataset included in this package, which contains a subset of the TCGA mRNA Ovarian serous cystadenocarcinoma data generated using Affymetrix HTHGU133a arrays. See `?tcgaov` for details about the data.  

In the example below we use the `r_cluster_data` to create the environment based clusters, and their summaries. We then use the `r_prepare_data` function to get it into proper form for regression routines such as `earth::earth`, `glmnet::cv.glmnet`, and `ncvreg::ncvreg`.

## Extract the relevant data

```{r, eval = TRUE}
# load the data
data("tcgaov")
tcgaov[1:5,1:6, with = FALSE]

# use log survival as the response
Y <- log(tcgaov[["OS"]])

# specify the environment variable
E <- tcgaov[["E"]]

# specify the matrix of genes only
genes <- as.matrix(tcgaov[,-c("OS","rn","subtype","E","status"),with = FALSE])

# for this example the training set will be all subjects.
# change `p` argument to create a train and test set.
trainIndex <- drop(caret::createDataPartition(Y, p = 1, list = FALSE, times = 1))
testIndex <- trainIndex
```


## Cluster the data and calculate cluster representations

We cluster the genes using the correlation matrix (specified by `cluster_distance = "corr"`) and the difference of the exposure dependent correlation matrices (specified by `eclust_distance = "diffcorr"`)

```{r}
cluster_res <- r_cluster_data(data = genes,
                              response = Y,
                              exposure = E,
                              train_index = trainIndex,
                              test_index = testIndex,
                              cluster_distance = "corr",
                              eclust_distance = "diffcorr",
                              measure_distance = "euclidean",
                              clustMethod = "hclust",
                              cutMethod = "dynamic",
                              method = "average",
                              nPC = 1,
                              minimum_cluster_size = 30)

# the number of clusters determined by the similarity matrices specified
# in the cluster_distance and eclust_distance arguments. This will always be larger
# than cluster_res$clustersAll$nclusters which is based on the similarity matrix
# specified in the cluster_distance argument
cluster_res$clustersAddon$nclusters

# the number of clusters determined by the similarity matrices specified
# in the cluster_distance argument only
cluster_res$clustersAll$nclusters

# what's in the cluster_res object
names(cluster_res)
```

## Prepare data for input in any regression routine

Now we use the `r_prepare_data` function, where we are using the average expression from each cluster as feaand their interaction with E as features in the regression model:

```{r}
# prepare data for use with earth function
avg_eclust_interaction <- r_prepare_data(
  data = cbind(cluster_res$clustersAddon$averageExpr, 
               Y = Y[trainIndex],
               E = E[trainIndex]),
  response = "Y", exposure = "E")

head(avg_eclust_interaction[["X"]])
```


## Fit a regression model

At this stage, you can decide which regression model to use. Here we choose the MARS model from the `earth` package, but you may choose regression models from any number of packages (e.g. see the [extensive list of models](https://topepo.github.io/caret/available-models.html) of models available in the `caret` package).

```{r}
fit_earth <- earth::earth(x = avg_eclust_interaction[["X"]], 
                          y = avg_eclust_interaction[["Y"]], 
                          pmethod = "backward", 
                          keepxy = TRUE, 
                          degree = 2, 
                          trace = 1, 
                          nk = 1000)

coef(fit_earth)
```

You can also install the `plotmo` package to visualise the relationships between the hinge functions and the response using `plotmo::plotmo(fit_earth)`. 

## Determine the features that have been selected

The `u_extract_selected_earth` is a utility function in this package to extract the selected predictors from the MARS model: 
```{r}
u_extract_selected_earth(fit_earth)
```

We that genes in clusters 1, 7 and 9 were selected. We also see that the interaction between the genes in cluster 5 and the environment was selected and has the highest variable importance. We can see the genes involved using the `cluster_res$clustersAddonMembership` object: 

```{r}
# Genes in cluster 5
cluster_res$clustersAddonMembership[cluster %in% 5]

# variable importance
earth::evimp(fit_earth)
```

